Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-28 paper/paper.tex
Back to 25 pages

\begin{lemma}[Wne application\footnote{\dotv{normalform.v}{wne\_app}}]
  \label{lemma:wnewn}
  If $[[wne a]]$ and $[[wn b]]$, then $[[wne (a b)]]$.
\end{lemma}
\begin{proof}
  Immediate by lexicographical induction over the length of the reduction sequences in
  $[[wne a]]$ and $[[wn b]]$.
\end{proof}

@@

The normalization property then follows as a corollary of the
fundamental theorem.
\begin{corollary}[Existence of $\beta$-normal forms\footnote{\dotv{soundnessopen.v}{mltt\_normalizing}}]
  \label{corollary:exbetanf}
  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
\end{corollary}
\begin{proof}
  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
  We pick the $[[rho]]$ to be the identity substitution $[[idtm]]$, which injects
  variables as terms. The side condition $[[idtm |=
  G]]$ is satisfied since Lemma~\ref{lemma:adequacy} says neutral terms,
  including variables, semantically inhabit any $[[S0]]$ where
  $[[S0]]$ is the interpretation of some type. With our choice of
  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
  $[[S]]$. By Lemma~\ref{lemma:adequacy}, we
  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
\end{proof}

@@

\Rref{I-PiCoq} shows how \rref{I-Pi} is actually encoded in our
mechanized proof.
\begin{center}
  \drule[width=5in]{I-PiCoq}
\end{center}
Compared to \rref{I-Pi}, \rref{I-PiCoq} replaces the function
$[[F]]$ with a total relation $[[R]]$. The equivalence of these two
rules follows from the fact that the logical relation is a
partial function (Lemma~\ref{lemma:logreldeter}). In set-theoretic
notation, \rref{I-Pi} is more readable. However, if we want to encode
the same rule in Coq, we must encode $[[F]]$ as a relation (with type
\texttt{tm -> (tm -> Prop) -> Prop}) that satisfies the following functionality constraint.
\begin{minted}{coq}
              forall a S0 S1, F a S0 -> F a S1 -> S0 = S1.
\end{minted}
In comparison, \rref{I-PiCoq} does not require this side condition and
results in a simpler definition.

@@

In contrast, \citet{decagda} and \citet{nbeincoq} employ a deterministic weak
head reduction relation. The use of a deterministic reduction relation makes
the functionality of a logical relation trivial to prove, but the downside is
that weak head reduction (denoted by $\leadsto^{wh}$) or leftmost-outermost
reduction (denoted by $\leadsto^{lo}$) does not satisfy even the weaker
substitution property (Lemma~\ref{lemma:parsubst}).  With this alternative
formulation, we would have to prove how the deterministic reduction relates to
a full non-deterministic reduction relation to prove the fundamental
theorem. This would amount to proving the factorization theorem of the
deterministic reduction relation with respect to full non-deterministic
reduction. Instead, using parallel reduction in \rref{I-Red} allows us to
delay the factorization property until we need to justify the use
leftmost-outermost reduction as a deterministic reduction strategy for
normalization.

>>>>

\begin{lemma}[Wne application\footnote{\dotv{normalform.v}{wne\_app}}]
  \label{lemma:wnewn}
  If $[[wne a]]$ and $[[wn b]]$, then $[[wne (a b)]]$.
\end{lemma}
\begin{proof}
  Immediate by induction over the length of the reduction sequences in
  $[[wne a]]$ and $[[wn b]]$.
\end{proof}

@@

The normalization property then follows as a corollary of the
fundamental theorem.
\begin{corollary}[Existence of $\beta$-normal forms\footnote{\dotv{soundnessopen.v}{mltt\_normalizing}}]
  \label{corollary:exbetanf}
  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
\end{corollary}
\begin{proof}
  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
  We pick the $[[rho]]$ to be the identity substitution $[[idtm]]$, which injects
  variables as terms. The side condition $[[idtm |=
  G]]$ is satisfied since Lemma~\ref{lemma:adequacy} says neutral terms,
  including variables, semantically inhabit any $[[S0]]$ where
  $[[S0]]$ is the interpretation of some type. With our choice of
  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
  $[[S]]$. By Lemma~\ref{lemma:adequacy}, we
  conclude that $[[wn a]]$ and $[[wn A]]$.
\end{proof}

@@

\Rref{I-PiCoq} shows how \rref{I-Pi} is actually encoded in our
mechanized proof.
\begin{center}
  \drule[width=5in]{I-PiCoq}
\end{center}
Compared to \rref{I-Pi}, \rref{I-PiCoq} replaces the function
$[[F]]$ with a total relation $[[R]]$. The equivalence of these two
rules follows from the fact that the logical relation is a
partial function (Lemma~\ref{lemma:logreldeter}). In set-theoretic
notation, \rref{I-Pi} is more readable. However, if we want to encode
the same rule in Coq, we must encode $[[F]]$ as a relation (with type
\texttt{tm -> (tm -> Prop) -> Prop}) that satisfies the functionality constraint:
\mintinline{coq}{forall a S0 S1, F a S0 -> F a S1 -> S0 = S1}.
In comparison, \rref{I-PiCoq} does not require this side condition and
results in a simpler definition.

@@

In contrast, \citet{decagda} and \citet{nbeincoq} employ a deterministic weak
head reduction relation. % The use of a
A deterministic reduction relation makes
the functionality of a logical relation trivial to prove, but fails to
satisfy the substitution property (Lemma~\ref{lemma:parsubst}), an
issue that has been observed by
\citet{casinghino:combining-proofs-programs}. If we had chosen to work
with a deterministic reduction relation, we would likely need
results such as the factorization theorem~\citep{takahashi-parallel-reduction,factorization-essentially} in our development before we
can prove the fundamental theorem, leading to a more complicated proof.
% With
% this alternative formulation, we would have to prove how the deterministic reduction relates to
% a full non-deterministic reduction relation to prove the fundamental
% theorem. This would amount to proving the factorization theorem of the
% deterministic reduction relation with respect to full non-deterministic
% reduction. Instead, using parallel reduction in \rref{I-Red} allows us to
% delay the factorization property until we need to justify the use
% leftmost-outermost reduction as a deterministic reduction strategy for
% normalization.
